Translation of B Implementations to the LLVM: Control and Modularity

### 2013-08-23 translation.tex
Added a new example with operators pred and succ and a scalar constant. Updated translate.py and translation.tex to reflect the changes necessary to process this example and simulate the generated LLVM code.

        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred}} \\

@@

\DD{Include \B{Vari} as one of the alternative for an \B{Expr}.}

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to LLVM constants. To avoid
  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

@@

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each constant} \B{cst} \textit{generate} \\
      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

@@

\item $\intf(\B{M})_c$ is the definition of the component constants. It maps
the constant names to their type:
$\intf(\B{M})_c \isdef \trad{\B{M.const}}{\ListOf{\B{Cons}}}$ and
$\trad{\B{c}}{\B{Cons}} \isdef \B{c} \mapsto \trad{\B{c.type}}{\B{Type}}$.

@@

The actual LLVM source code is generated following the template shown in
figure~\ref{fig:skel} (a), as follows:
\begin{align*}
  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
  \quad \PH{\trad{\B{i}}{State}}} \\
& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
& \llvm{\PH{\trad{\B{i.init}}{Init}} 
  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
We shall now define each rule referenced in this template.
\begin{enumerate}
\item For the type definitions corresponding to the state of the imported
  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
    type \PH{\intf(\B{i})_t}\nl"}$.
\item For the declarations of the initialization functions in the imported
  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
\item For the definitions of the imported constants: $\trad{\B{i}, \intf}{ICons}
  \isdef \trad{$\intf(\B{i})_c$}{\ListOf{Cons}}$, where
  \noindent $\trad{$\B{c} \mapsto t)$}{Cons} \isdef \llvm{"\PH{\Global(\B{c})} =
    external constant \PH{t}\nl"}$.
\item For the declarations of the imported operations: $\trad{\B{i},
    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
\item The translation of constant definitions is presented in section~\ref{sec:data}.
\item For the definition of the type for component states: \\
\noindent$\trad{\B{M}}{State} \isdef
\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
\item For the initialization, the definition of a function is generated. It
  first initializes the imported modules and then the current module (see
  section~\ref{sec:control}).
\item For each operation, a function definition is generated (see
  section~\ref{sec:control}).
\end{enumerate}
Let $\idx$ be a function that maps each imported machine and state variable to
its position (starting from zero).
\section{Translation: Data}
\label{sec:data}

@@

B constants are translated to constants in the LLVM program, as described in the
following rule, where parameter \B{c} denote the constant, $M$ the name of the
module and $\Global$ is the mapping between source and target identifiers. Two
results are produced: the LLVM source code and an updated identifier mapping:
\begin{small}
\begin{align*}
  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
\end{align*}
\end{small}

The definition of the translation for expressions is described next in
section~\ref{sec:expr}.

\section{Translation: Expressions
  \label{sec:expr}}

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

@@

Next is the translation of the unary \B{succ} operator:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \IN \LET w = \variable \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows: \DD{Rewrite so
  that the temporary variable names are generated in the correct order.}

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
& \ELSE \LET \B{i}, \B{il'} = \B{il}  \AND p = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
& \quad \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
\end{align*}

@@

>>>>

        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Vari}} \\

@@

@@

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to their value in the
  LLVM platform.
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

@@

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

@@

@@

The actual LLVM source code is generated following the template shown in
figure~\ref{fig:skel} (a), as follows:
\begin{align*}
  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
  \quad \PH{\trad{\B{i}}{State}}} \\
& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
& \llvm{\PH{\trad{\B{i.init}}{Init}} 
  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
We shall now define each rule referenced in this template.
\begin{enumerate}
\item For the type definitions corresponding to the state of the imported
  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
    type \PH{\intf(\B{i})_t}\nl"}$.
\item For the declarations of the initialization functions in the imported
  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
\item For the declarations of the imported operations: $\trad{\B{i},
    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
\item The translation of constant definitions is presented in section~\ref{sec:data}.
\item For the definition of the type for component states: \\
\noindent$\trad{\B{M}}{State} \isdef
\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
\item For the initialization, the definition of a function is generated. It
  first initializes the imported modules and then the current module (see
  section~\ref{sec:control}).
\item For each operation, a function definition is generated (see
  section~\ref{sec:control}).
\end{enumerate}
Let $\idx$ be a function that maps each imported machine and state variable to
its position (starting from zero).
\section{Translation: Data}
\label{sec:data}

@@

B constants shall be translated to constants in the LLVM program. The
translation of scalar constants consists in substituting them with their value.
\DD{The translation of non-scalar constants will be defined in a next version of
  this document.}  The definition of the translation for expressions is
described next in section~\ref{sec:expr}.

\section{Translation: Expressions}
\label{sec:expr}

@@

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

@@

Next is the translation of the unary \B{succ} operator:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t = \trad{n.arg}{\B{expr}} \IN  \\
  & \quad \LET w = \variable \IN \\
  & \quad \quad \llvm{"\PH{p}} \\
  & \quad \quad \llvm{\PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

@@

One issue in the translation of instructions is the disruption of the flow graph
intro a linear sequence of instructions and branches. This requires the creation
of instruction labels. As a consequence, our definition of the translation of
instructions has two kinds of rules, those that take a label as parameter, and
those that do not. Such label parameter indicates where the control should go
after the corresponding execution is executed.

\begin{align*}
\trad{\B{i}, $\ell$}{\B{Inst}}_L \isdef
& \IF \B{i} \mbox{ is \B{If}} \THEN \trad{\B{i}, $\ell$}{\B{If}}_L \\
& \ELSIF \B{i} \mbox{ is \B{While}} \THEN \trad{\B{i}, $\ell$}{\B{While}}_L \\
& \ELSIF \B{i} \mbox{ is \B{Blk}} \THEN \trad{\B{i.body}, $\ell$}{\ListOf{\B{Inst}}}_L \\
& \ELSE \LET p = \trad{\B{i}}{\B{Inst}} \IN \\
& \quad \llvm{"\PH{p}} \\
& \quad \llvm{ br label \%\PH{\ell} \nl"}
\end{align*}

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows:

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
& \ELSE \\
& \LET \B{i}, \B{il'} = \B{il}  \IN \\
& \quad \quad \LET p_1 = 
\begin{array}[t]{l}
  \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \\
  \quad \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
  \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"} \IN
\end{array} \\
& \quad \quad \IN \\
& \quad \quad \quad \LET p_2 = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \\
& \quad \quad \quad \quad \llvm{"\PH{p_1} \PH{p_2}"}
\end{align*}

@@

\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|}
\hline
Example nº & 01 & 02 \\
\hline
variables & \yes & \yes \\
constants & \no & \yes \\
op. input & \yes & \no \\
op. output & \yes & \no \\
\hline
\multicolumn{3}{|c|}{instructions} \\
\hline
block & \yes & \no \\
local var & \no & \no \\
identity & \no & \no \\
becomes equal & \yes & \yes \\
operation call & \no & \no \\
condition & \yes & \no \\
case & \no & \no \\
assertion & \no & \no \\
sequence & \yes & \yes \\
while & \no & \no \\
\hline
\multicolumn{3}{|c|}{expressions} \\
\hline
succ & \no & \yes \\
pred & \no & \yes \\
+ & \yes & \no \\
- & \yes & \no \\
\hline
\multicolumn{3}{|c|}{formulas} \\
\hline
and & \no & \no \\
or & \no & \no \\
not & \no & \no \\
comparison & \yes & \yes \\
\hline
\end{tabular}
\end{center}
